---
title: "Colour, Fill, and Shape Palettes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colour, Fill, and Shape Palettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(ggplot2)
library(ggprism)
library(patchwork)
```

`ggprism` includes colour and fill scales with various palettes that mirror
the colour schemes available in GraphPad Prism. 

**Disclaimer:** some of these palettes are
quite nice, some are pretty ugly and not useful, and most are not colour blind 
friendly as far as I know (except `palette = "colorblind_safe"`). It would 
probably be safer to use other colour scales e.g. those in `RColorBrewer` for 
any serious application or work to be published. 

In any case, these scales are used to colour the 
data elements of ggplots and can be used separately or together with 
`theme_prism()`. Additionally, a shape scale with 3 palettes is included which
is similar to the shapes available in Prism. This vignette outlines how to
use `scale_colour_prism()` (or `scale_color_prism()`), `scale_fill_prism()`,
and `scale_shape_prism()`.

## Colour scales

First we'll create a base plot.

```{r, fig.width=3.6, fig.asp=0.9}
# create a base plot to compare colour scales
base <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(colour = factor(cyl), shape = factor(cyl)), size = 3) + 
  theme_prism() + 
  theme(legend.position = c(0.8, 0.8))

base
```

`scale_colour_prism()` (US spelling also works) functions pretty much 
identically to `scale_colour_manual()` from `ggplot2`. The default palette for
`scale_colour_prism()` is `"colors"`.

```{r, fig.width=7.2, fig.asp=0.5}
# compare manual colour scale with prism colour scale
p1 <- base + scale_colour_manual(values = c("blue", "red", "green3"))
p2 <- base + scale_colour_prism()

p1 + p2
```

There are several different colour palettes that are built into 
`scale_colour_prism()`.

```{r}
# see names and lengths of available scale_colour_prism() palettes
lengths(ggprism_data$colour_palettes)
```

Using the palettes is simple.

```{r, fig.width=7.2, fig.asp=0.5}
# try out some different colour palettes
p1 <- base + scale_colour_prism(palette = "purple_passion")
p2 <- base + scale_colour_prism(palette = "candy_bright")

p1 + p2
```

You can preview a single palette using the `preview_theme()` function. It shows 
a sample plot with the colour scale applied (as well as `scale_fill_prism()` and
`theme_prism()`). See 
[this page](https://csdaw.github.io/ggprism/articles/web-only/palettes.html) for
images of most available colour palettes.

```{r, fig.width=4.5}
preview_theme("flames")
```

## Fill scales

First we'll create a base plot.

```{r, fig.width=3.6, fig.asp=0.9}
# create a base plot to compare fill scales
base <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(fill = factor(cyl), shape = factor(cyl)), size = 3) + 
  theme_prism() + 
  theme(legend.position = c(0.8, 0.8)) + 
  scale_shape_prism(palette = "filled")

base
```

`scale_fill_prism()` functions pretty much 
identically to `scale_fill_manual()` from `ggplot2`. The default palette for
`scale_fill_prism()` is `"colors"`.

```{r, fig.width=7.2, fig.asp=0.5}
# compare manual fill scale with prism fill scale
p1 <- base + scale_fill_manual(values = c("blue", "red", "green3"))
p2 <- base + scale_fill_prism()

p1 + p2
```

There are several different fill palettes that are built into 
`scale_fill_prism()`.

```{r}
# see names and lengths of available scale_fill_prism() palettes
lengths(ggprism_data$fill_palettes)
```

Using the palettes is simple.

```{r, fig.width=7.2, fig.asp=0.5}
# try out some different fill palettes
p1 <- base + scale_fill_prism(palette = "colorblind_safe")
p2 <- base + scale_fill_prism(palette = "neon")

p1 + p2
```

You can preview a single palette using the `preview_theme()` function. It shows 
a sample plot with the colour scale applied (as well as `scale_colour_prism()` and
`theme_prism()`). See 
[this page](https://csdaw.github.io/ggprism/articles/web-only/palettes.html) for
images of most available fill palettes.

```{r, fig.width=4.5}
preview_theme("diazo")
```

## Shape scales

The order of shapes used in GraphPad Prism is slightly different to `ggplot2`,
hence 3 different shape palettes are included in the `ggprism` function
`scale_shape_prism()`.

We can see the names and lengths of these shape palettes. Naturally, 
`"default"` is the default palette and does not include any shape with a fill. 
The `"filled"` palette should be used when you also want to use a fill scale, 
as the first 5 symbols have a fill. The `"complete"` palette symbols 5-9 have
a fill.

```{r}
# see names and lengths of available scale_shape_prism() palettes
lapply(ggprism_data$shape_palettes, nrow)
```

We can also define a function to see all the shapes in a specific palette.

```{r, fig.width=3.6, fig.asp=0.9}
# define a function for convenience
show_shapes <- function(palette) {
  df_shapes <- ggprism_data$shape_palettes[[palette]][, -1]
  df_shapes$pch_f <- factor(df_shapes$pch, levels = df_shapes$pch)

  ggplot(df_shapes, aes(x = 0, y = 0, shape = pch)) +
    geom_point(aes(shape = pch), size = 5, fill = 'red') +
    scale_shape_identity() +
    facet_wrap(~ pch_f) +
    theme_void()
}

# show the shapes in the palette "complete"
show_shapes("complete")
```

We'll define a base plot to see `scale_shape_prism()` in action.

```{r, fig.width=3.6, fig.asp=0.9}
# create a base plot to compare shape scales
base <- ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(aes(colour = class, fill = class, shape = class)) + 
  theme_prism(base_size = 11, base_fontface = "plain", border = TRUE) +
  theme(legend.position = c(0.8, 0.75),
        plot.subtitle = element_text(face = "bold"),
        legend.key.height = unit(10, "pt")) +
  coord_cartesian(clip = "off") + 
  scale_colour_prism(palette = "floral") + 
  scale_fill_prism(palette = "floral")

base
```

Then we'll compare the 3 shape palettes to the `ggplot2` base plot.

```{r, fig.width=7, fig.height=6}
# compare shape scales
p1 <- base
p2 <- base + scale_shape_prism(palette = "default") + 
  labs(subtitle = "default")
p3 <- base + scale_shape_prism(palette = "filled") + 
  labs(subtitle = "filled")
p4 <- base + scale_shape_prism(palette = "complete") + 
  labs(subtitle = "complete")

(p1 + p2) / (p3 + p4)
```
