---
title: "Adding p-values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding p-values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(ggplot2)
library(ggprism)

library(dplyr)
library(tidyr)
library(rstatix)
```

Arguably one of the most popular features of GraphPad Prism is adding p-values
to plots. Indeed in [Prism 9](https://www.graphpad.com/support/faq/prism-900-release-notes/), 
GraphPad have added a feature to automatically
perform pairwise comparisons and add the resulting p-values with brackets to
the graph.

`ggprism` includes the `add_pvalue()` function to add p-values with or without
brackets to ggplots. This vignette will go through the many ways in which this
function can be used.

This function is a re-written version of `stat_pvalue_manual()` from the 
[`ggpubr` package](https://github.com/kassambara/ggpubr), which itself is 
based on the `geom_signif()` function from the 
[`ggsignif` package](https://github.com/const-ae/ggsignif). Compared to
`stat_pvalue_manual()`, the `add_pvalue()` function is easier to use, is more
robust with less dependencies, and has more customisable brackets.

## Basic use (brackets)

## Basic use (no brackets)

## Change the look

## Using the `rstatix` package

## Many more examples

### Compare two means

### Compare mean vs reference

### Multiple pairwise comparisons

### Grouped pairwise comparisons

### Facets
