---
title: "Adding p-values"
output:
  rmarkdown::html_vignette:
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Adding p-values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(ggplot2)
library(ggprism)
library(patchwork)
library(magrittr)
```

Arguably one of the most popular features of GraphPad Prism is adding p-values
to plots. Indeed in [Prism 9](https://www.graphpad.com/updates/prism-900-release-notes), 
GraphPad have added a feature to automatically
perform pairwise comparisons and add the resulting p-values with brackets to
the graph.

`ggprism` includes the `add_pvalue()` function to add p-values with or without
brackets to ggplots. This vignette will go through the many ways in which this
function can be used.

This function is a re-written version of `stat_pvalue_manual()` from the 
[`ggpubr` package](https://github.com/kassambara/ggpubr), which itself is 
based on the `geom_signif()` function from the 
[`ggsignif` package](https://github.com/const-ae/ggsignif). Compared to
`stat_pvalue_manual()`, the `add_pvalue()` function is: easier to use, more
robust with less dependencies, and has more customisable brackets.

## Basic use (brackets)

To add significance brackets to a plot, you need a minimal data.frame with 4
columns and a number of rows corresponding to the number of brackets you want
to add. The 4 columns should correspond to these 4 function arguments:

1. xmin, the left hand position of the bracket (default column name is `"group1"`)
1. xmax, the right hand position of the bracket (default column name is `"group2"`)
1. label, the text to go on the brackets (default column name is `"label"`)
1. y.position, the vertical position of the bracket (default column name is `"y.position"`)

For grouped or faceted data you'll also need a column which is named according
to the grouping variable. See the **Many more examples** section for help with
this/examples.

Let's see how this works in practice. First we'll plot the `sleep` data set.

```{r}
str(sleep)
```


```{r, fig.width=4, fig.height=3.5}
# create a jitter plot of the sleep data set
# and indicate the means
p <- ggplot(sleep, aes(x = group, y = extra)) +
  geom_jitter(aes(shape = group), width = 0.1) + 
  stat_summary(geom = "crossbar", fun = mean, colour = "red", width = 0.2) + 
  theme_prism() + 
  theme(legend.position = "none")
p
```

Next we'll perform a t-test and obtain a p-value for the difference between
the two means.

```{r}
# perform a t-test and obtain the p-value
result <- t.test(extra ~ group, data = sleep)$p.value
result <- signif(result, digits = 3)
result
```

Now we'll construct a p-value data.frame for `add_pvalue()` to use.

```{r}
df_p_val <- data.frame(
  group1 = "1",
  group2 = "2",
  label = result,
  y.position = 6
)
```

And finally we'll add this p-value to our plot. Because we have used the 
default column names (see above) in our p-value table we don't necessarily 
have to specify any arguments of `add_pvalue()`. However, here we'll do it
for clarity's sake. Additionally, if your p-value table has special 
column names, you will need to specify them in `add_pvalue()`.

```{r, fig.height=3.5}
# add p-value brackets
p1 <- p + add_pvalue(df_p_val,
                     xmin = "group1",
                     xmax = "group2",
                     label = "label",
                     y.position = "y.position") 

# change column names to something silly
colnames(df_p_val) <- c("apple", "banana", "some_label", "some_y_position")

# add p-value brackets again
p2 <- p + add_pvalue(df_p_val,
                     xmin = "apple",
                     xmax = "banana",
                     label = "some_label",
                     y.position = "some_y_position")

p1 + p2
```

```{r}
# return column names back to default
colnames(df_p_val) <- c("group1", "group2", "label", "y.position")
```

You can easily change how the bracket and label looks. You can make the label 
a `glue` expression. You can also change the tip length of the bracket.
Lastly, you can flip the label when using `coord_flip()`.

```{r, fig.width=7, fig.height=7}
# change bracket and label aesthetics
p1 <- p + add_pvalue(df_p_val,
                     colour = "red", # label
                     label.size = 8, # label
                     fontface = "bold", # label
                     fontfamily = "serif", # label
                     angle = 45, # label
                     hjust = 1, # label
                     vjust = 2, # label
                     bracket.colour = "blue", # bracket
                     bracket.size = 1, # bracket
                     linetype = "dashed", # bracket
                     lineend = "round") # bracket

# use glue expression for label
p2 <- p + add_pvalue(df_p_val, label = "p = {label}")

# make bracket tips longer and use coord_flip
p3 <- p + add_pvalue(df_p_val, tip.length = 0.1) + 
  coord_flip()

# change bracket tips independently
# (make one side disappear and the other longer)
p4 <- p + add_pvalue(df_p_val, tip.length = c(0.1, 0))

(p1 + p2) / (p3 + p4)
```

## Basic use (no brackets)

Even if you don't want brackets, `add_pvalue()` is also useful for adding
significance text to plots with the correct/automatic positioning.

In the example above, if you wanted the text but not the bracket, you can just
use the `remove.bracket` argument. In this case, you must use the `x` argument
to change the x position of the text.

```{r, fig.height=3.5}
# position label above "group1"
p1 <- p + add_pvalue(df_p_val, label = "p = {label}", 
                     remove.bracket = TRUE, x = 1)

# position label between x = 1 and x = 2
p2 <- p + add_pvalue(df_p_val, label = "p = {label}", 
                     remove.bracket = TRUE, x = 1.5)

p1 + p2
```

Here is another example of 'text only' plot using the `ToothGrowth` data set.

```{r}
str(ToothGrowth)
```

```{r, fig.width=4, fig.height=3.5}
# create a box plot of the ToothGrowth data set
p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) +
  geom_boxplot(aes(fill = dose), colour = "black") + 
  theme_prism() + 
  theme(legend.position = "none")
p
```

Next we'll perform two t-tests and compare the means against `dose = 0.5` as
a reference group. Then we'll correct the p-values for multiple testing.

```{r}
# compare means again reference
result1 <- t.test(len ~ dose, 
                  data = subset(ToothGrowth, dose %in% c(0.5, 1.0)))$p.value
result2 <- t.test(len ~ dose, 
                  data = subset(ToothGrowth, dose %in% c(0.5, 2.0)))$p.value

# Benjamini-Hochberg correction for multiple testing
result <- p.adjust(c(result1, result2), method = "BH")
```

We can now construct a p-value table. Note that in this case we don't need to
to specify a `"group2"` column for xmax. This is because text-only p-value
annotations just have an x position (`x`) and not an x range (`xmin` and `xmax`).

```{r}
# don't need group2 column (i.e. xmax)
# instead just specify x position in the same way as y.position
df_p_val <- data.frame(
  group1 = c(0.5, 0.5),
  group2 = c(1, 2),
  x = c(2, 3),
  label = signif(result, digits = 3),
  y.position = c(35, 35)
)
```

Then we add the p-values to the plot. As before, you can change how the labels
look quite easily.

```{r, fig.height=3.5}
p1 <- p + add_pvalue(df_p_val, 
                     xmin = "group1", 
                     x = "x", 
                     label = "label",
                     y.position = "y.position")

p2 <- p + add_pvalue(df_p_val, 
                     xmin = "group1", 
                     x = "x", 
                     label = "p = {label}",
                     y.position = "y.position",
                     label.size = 3.2,
                     fontface = "bold")

p1 + p2
```

If you want the label number format to look nicer you can provide a column name to
`label` with _plotmath_ expressions and set `parse = TRUE`. This works with or
without brackets.

```{r, fig.height=3.5}
# plotmath expression to have superscript exponent
df_p_val$p.exprs <- paste0("P==1*x*10^", round(log10(df_p_val$label), 0))

# as above but with italics
df_p_val$p.exprs.ital <- lapply(
  paste(round(log10(df_p_val$label), 0)),
  function(x) bquote(italic("P = 1x10"^.(x)))
)

p1 <- p + add_pvalue(df_p_val, 
                     xmin = "group1", 
                     x = "x", 
                     label = "p.exprs",
                     y.position = "y.position",
                     parse = TRUE)

p2 <- p + add_pvalue(df_p_val, 
                     xmin = "group1", 
                     x = "x", 
                     label = "p.exprs.ital",
                     y.position = "y.position",
                     parse = TRUE)

p1 + p2
```


## Using the `rstatix` package

As `add_pvalue()` is ultimately just a rewritten version of 
`stat_pvalue_manual()`, it works well with the 
[`rstatix` package](https://github.com/kassambara/rstatix).

With [`rstatix`](https://github.com/kassambara/rstatix), you can perform the
statistical test and create the p-value table with the appropriate
x and y position automatically, in a single step.

```{r, fig.width=4, fig.height=3.5}
df_p_val <- rstatix::t_test(ToothGrowth, len ~ dose, ref.group = "0.5") %>% 
  rstatix::add_xy_position()

p + add_pvalue(df_p_val, 
               label = "p = {p.adj}",
               remove.bracket = TRUE)
```

## Many more examples

Here we will use `add_pvalue()` and 
[`rstatix`](https://github.com/kassambara/rstatix) to show many more examples
of how to add p-values to different plots.

### Compare two means

Compare mean `len` depending on `supp`. Error bars indicate 
1 standard deviation from the mean.

```{r, fig.width=4, fig.height=3.5}
df_p_val <- rstatix::t_test(ToothGrowth, len ~ supp) %>% 
  rstatix::add_x_position()

p <- ggplot(ToothGrowth, aes(x = factor(supp), y = len)) + 
  stat_summary(geom = "col", fun = mean) + 
  stat_summary(geom = "errorbar", 
               fun = mean,
               fun.min = function(x) mean(x) - sd(x),
               fun.max = function(x) mean(x) + sd(x),
               width = 0.3) + 
  theme_prism() + 
  coord_cartesian(ylim = c(0, 35)) + 
  scale_y_continuous(breaks = seq(0, 35, 5), expand = c(0, 0))

# normal plot
p + add_pvalue(df_p_val, y.position = 30)
```

### Compare mean vs reference

Compare mean `len` of each dose to `dose = 0.5`. Error bars indicate 
1 standard deviation from the mean.

```{r, fig.height=3.5}
df_p_val <- rstatix::t_test(ToothGrowth, len ~ dose, ref.group = "0.5") %>% 
  rstatix::add_xy_position()

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  stat_summary(geom = "col", fun = mean) + 
  stat_summary(geom = "errorbar", 
               fun = mean,
               fun.min = function(x) mean(x) - sd(x),
               fun.max = function(x) mean(x) + sd(x),
               width = 0.3) + 
  theme_prism() + 
  coord_cartesian(ylim = c(0, 40)) + 
  scale_y_continuous(breaks = seq(0, 40, 5), expand = c(0, 0))

# with brackets
p1 <- p + add_pvalue(df_p_val, label = "p.adj.signif")

# without brackets
p2 <- p + add_pvalue(df_p_val, label = "p.adj.signif", remove.bracket = TRUE)

p1 + p2
```

Now, compare overall mean `len` (base mean) to the mean `len` for each `dose`. 
Error bars indicate 1 standard deviation from the mean.

```{r, fig.width=4, fig.height=3.5}
df_p_val <- rstatix::t_test(ToothGrowth, len ~ dose, ref.group = "all")

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  stat_summary(geom = "col", fun = mean) + 
  stat_summary(geom = "errorbar", 
               fun = mean,
               fun.min = function(x) mean(x) - sd(x),
               fun.max = function(x) mean(x) + sd(x),
               width = 0.3) + 
  theme_prism() + 
  coord_cartesian(ylim = c(0, 40)) + 
  scale_y_continuous(breaks = seq(0, 40, 5), expand = c(0, 0))

p + add_pvalue(df_p_val, 
               label = "p.adj.signif", 
               y.position = 35)
```

Now, compare the mean `len` for each `dose` to some arbitrary value, say
`26` in this case. Error bars indicate 1 standard deviation from the mean.

```{r, fig.width=4, fig.height=3.5}
df_p_val <- ToothGrowth %>% 
  rstatix::group_by(factor(dose)) %>% 
  rstatix::t_test(len ~ 1, mu = 26) %>% 
  rstatix::adjust_pvalue(p.col = "p", method = "holm") %>% 
  rstatix::add_significance(p.col = "p.adj")

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  stat_summary(geom = "col", fun = mean) + 
  stat_summary(geom = "errorbar", 
               fun = mean,
               fun.min = function(x) mean(x) - sd(x),
               fun.max = function(x) mean(x) + sd(x),
               width = 0.3) + 
  theme_prism() + 
  coord_cartesian(ylim = c(0, 40)) + 
  scale_y_continuous(breaks = seq(0, 40, 5), expand = c(0, 0))

# remember xmin and x are referring to the column dames in df_p_val
p + add_pvalue(df_p_val, 
               xmin = "group1", 
               x = "factor(dose)", 
               y = 37,
               label = "p.adj.signif")
```

### Multiple pairwise comparisons

Compare mean `len` across the 3 different `dose`. Use the `bracket.shorten`
argument to slightly shorten side-by-side brackets.

```{r, fig.width=4, fig.height=3.5}
df_p_val <- rstatix::t_test(ToothGrowth, len ~ dose)

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  geom_violin(trim = FALSE) + 
  geom_boxplot(width = 0.2) +
  theme_prism() + 
  coord_cartesian(ylim = c(0, 47)) + 
  scale_y_continuous(breaks = seq(0, 45, 5), expand = c(0, 0))

p + add_pvalue(df_p_val, 
               y.position = c(45, 41, 45),
               bracket.shorten = c(0.025, 0, 0.025))
```


### Grouped pairwise comparisons

Pairwise comparisons _between groups_ of the `ToothGrowth` data set, 
grouped according to `supp`. The boxplots and the brackets are automatically 
coloured according to `supp`. Three important points for this graph:

1. The p-value data.frame **must** have a column named with the grouping
variable (in this case named `supp`) and the column must contain the groups to
group by (in this case `"OJ" or "VC"`.
1. Any grouping of data **must** be done in the individual Geom functions 
(in this case `geom_boxplot()`) and **not** in the `ggplot()` function.
1. You can use the `step.group.by = "supp"` to automatically change the 
bracket spacing between different groups.

```{r, fig.width=5, fig.height=3.5}
df_p_val <- ToothGrowth %>% 
  rstatix::group_by(supp) %>% 
  rstatix::t_test(len ~ dose) %>% 
  rstatix::add_xy_position()

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  geom_boxplot(aes(fill = supp)) + 
  theme_prism()

# remember colour and step.group.by are referring to a column name in df_p_val
p + add_pvalue(df_p_val, 
               label = "p = {p.adj}",
               colour = "supp",
               fontface = "bold",
               step.group.by = "supp",
               step.increase = 0.1,
               tip.length = 0,
               bracket.colour = "black",
               show.legend = FALSE)
```

Pairwise comparisons _within groups_ of the `ToothGrowth` data set, 
grouped according to `supp`.

```{r, fig.width=5, fig.height=3.5}
df_p_val <- ToothGrowth %>%
  rstatix::group_by(dose) %>%
  rstatix::t_test(len ~ supp) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "bonferroni") %>%
  rstatix::add_significance(p.col = "p.adj") %>% 
  rstatix::add_xy_position(x = "dose", dodge = 0.8) # important for positioning!

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  geom_boxplot(aes(fill = supp)) + 
  theme_prism() + 
  coord_cartesian(ylim = c(0, 40))

p + add_pvalue(df_p_val, 
               xmin = "xmin", 
               xmax = "xmax",
               label = "p = {p.adj}",
               tip.length = 0)
```

Pairwise comparisons _within groups_ and _between groups_ of the `ToothGrowth` 
data set, grouped according to `supp`. You can use `bracket.nudge.y` to 
slightly adjust the overall y position of the brackets instead of having to
redefine `df_p_val2`.

```{r, fig.width=5, fig.height=4}
df_p_val1 <- ToothGrowth %>%
  rstatix::group_by(dose) %>%
  rstatix::t_test(len ~ supp) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "bonferroni") %>%
  rstatix::add_significance(p.col = "p.adj") %>% 
  rstatix::add_xy_position(x = "dose", dodge = 0.8) # important for positioning!

df_p_val2 <- rstatix::t_test(ToothGrowth, len ~ dose, 
                             p.adjust.method = "bonferroni") %>% 
  rstatix::add_xy_position()

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  geom_boxplot(aes(fill = supp)) + 
  theme_prism() + 
  coord_cartesian(ylim = c(0, 45))

p + add_pvalue(df_p_val1, 
               xmin = "xmin", 
               xmax = "xmax",
               label = "p = {p.adj}",
               tip.length = 0) + 
  add_pvalue(df_p_val2,
             label = "p = {p.adj}",
             tip.length = 0.01,
             bracket.nudge.y = 2,
             step.increase = 0.015)
```

### Facets

Facet according to `dose` and then compare mean `len` between either `supp`.
It is important that the p-value table **must** have a column with the same 
name as the faceting variable (in this case `"dose"`).

```{r, fig.height=3.5}
df_p_val <- ToothGrowth %>% 
  rstatix::group_by(dose) %>% 
  rstatix::t_test(len ~ supp) %>% 
  rstatix::add_xy_position()

p <- ggplot(ToothGrowth, aes(x = factor(supp), y = len)) + 
  geom_boxplot(width = 0.2) + 
  facet_wrap(
    ~ dose, scales = "free", 
    labeller = labeller(dose = function(x) paste("dose =", x))
  ) + 
  theme_prism()

p + add_pvalue(df_p_val)
```

Facet according to `supp` and then compare mean `len` between the three `dose`.
It is important that the p-value table **must** have a column with the same 
name as the faceting variable (in this case `"supp"`).

```{r, fig.height=3.5}
df_p_val <- ToothGrowth %>% 
  rstatix::group_by(supp) %>% 
  rstatix::t_test(len ~ dose) %>% 
  rstatix::add_xy_position()

p <- ggplot(ToothGrowth, aes(x = factor(dose), y = len)) + 
  geom_boxplot(width = 0.4) + 
  facet_wrap(~ supp, scales = "free") + 
  theme_prism()

p + add_pvalue(df_p_val)
```

Facet according to some particular grouping variable called `grp` and `dose`, 
and then compare mean `len` between either `supp`. It is important that the 
p-value table **must** have columns with the same 
names as the two faceting variables (in this case `"grp"` and `"dose"`).

```{r, fig.height=7}
# add a grouping variable to ToothGrowth
tg <- ToothGrowth
tg$dose <- factor(tg$dose)
tg$grp <- factor(rep(c("grp1", "grp2"), 30))

# construct the p-value table by hand
df_p_val <- data.frame(
  group1 = c("OJ", "OJ"),
  group2 = c("VC", "VC"),
  p.adj = c(0.0449, 0.00265),
  y.position = c(22, 27),
  grp = c("grp1", "grp2"),
  dose = c("0.5", "1")
)

p <- ggplot(tg, aes(x = factor(supp), y = len)) + 
  geom_boxplot(width = 0.4) + 
  facet_wrap(grp ~ dose, scales = "free") + 
  theme_prism()

p + add_pvalue(df_p_val, bracket.nudge.y = 3)
```

