---
title: "Themes"
output:
  rmarkdown::html_vignette:
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(ggplot2)
library(ggprism)
library(patchwork)
```

`theme_prism()` is arguably the most useful part of this package. It is the 
main tool which is used make ggplots look like they were made in GraphPad Prism.
This vignette goes through of the options included in the function, and also
shows you how to make your own theme.

## Using `theme_prism()`

First we'll just define a base plot to use for the rest of the vignette.

```{r}
# define a base plot
base <- ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(aes(colour = class))
```

Then we'll apply `theme_prism()` and see how it looks just with the default
settings. By default the:

- Fontface for all text is bold except the legend text
- Font size is 14 pt
- Legend title is hidden (by setting `legend.title = element_blank()`)
- Background is white and all text and lines are black
- Default palette is `"black_and_white"`

You can easily bring back the legend title by overriding the `legend.title`
element.

```{r, fig.height=3.5}
# apply default theme
p1 <- base + 
  theme_prism() + 
  guides(colour = guide_legend(position = "inside")) +
  theme(legend.position.inside = c(0.8, 0.75),
        legend.key.height = unit(10, "pt"))
p2 <- base + 
  theme_prism() + 
  guides(colour = guide_legend(position = "inside")) +
  theme(legend.position.inside = c(0.8, 0.75),
        legend.key.height = unit(10, "pt"),
        legend.title = element_text())

p1 + p2
```

The overall theme text size is adjusted with the `base_size` 
argument, as with the standard ggplot2 themes such as `theme_bw()`. 

```{r, fig.height=3.5}
# redefine base plot without a legend for convenience
base <- ggplot(mpg, aes(x = displ, y = cty)) +
  geom_point(aes(colour = class), show.legend = FALSE)

# adjust overall theme size
p1 <- base + theme_prism(base_size = 10)
p2 <- base + theme_prism(base_size = 16)

p1 + p2
```

By default the `base_line_size` and `base_rect_size` arguments scales with the
`base_size`, but these can be overridden.

```{r, fig.height=3.5}
# adjust overall theme size with specific line size
p1 <- base + theme_prism(base_size = 14)
p2 <- base + theme_prism(base_size = 14, base_line_size = 0.2)

p1 + p2
```



You can also change the `base_fontface` (e.g. bold, plain, italic) and 
`base_family` (crossplatform compatibility fonts e.g. serif, sans, mono, 
or specific fonts e.g. Arial).

```{r, fig.height=3.5}
# change fontface or font family
p1 <- base + theme_prism(base_fontface = "plain")
p2 <- base + theme_prism(base_family = "mono")

p1 + p2
```

`theme_prism()` includes the `axis_text_angle` argument which is a convenient
way to rotate the x axis text (allowed angles are: 0, 45, 90, or 270).

```{r, fig.height=3.5}
# change x axis text angle
p1 <- base + theme_prism(axis_text_angle = 45)
p2 <- base + theme_prism(axis_text_angle = 90)

p1 + p2
```

Lastly, you can put a square border around the plot using the `border` argument.
Unfortunately, you need to turn of clipping in order for this to work/appear 
correct.

```{r, fig.height=3.5}
# add a border and adjust its thickness
p1 <- base + theme_prism(border = TRUE) + 
  coord_cartesian(clip = "off")
p2 <- base + theme_prism(border = TRUE, base_rect_size = 2) + # adjust thickness
  coord_cartesian(clip = "off")

p1 + p2
```

## Theme palettes

There are several different colour palettes that are built into `theme_prism()`.

```{r}
# see names of available theme_prism() palettes
names(ggprism_data$themes)
```

Using the palettes is simple.

```{r, fig.height=3.5}
# try out some different theme palettes
p1 <- base + theme_prism(palette = "purple_passion")
p2 <- base + theme_prism(palette = "candy_bright")

p1 + p2
```

You can preview a single palette using the `preview_theme()` function. It shows 
a sample plot with the theme applied (as well as `scale_colour_prism()` and
`scale_fill_prism()`). See 
[this page](https://csdaw.github.io/ggprism/articles/web-only/palettes.html) for
images of all 43 available `theme_prism()` palettes.

```{r, fig.width=4.5}
preview_theme("flames")
```

Note that several of the palettes happen to be identical. This is not a bug
but just how they were designed in GraphPad Prism.

```{r, fig.height=3.5}
# compare two identical theme palettes
p1 <- base + theme_prism(palette = "black_and_white")
p2 <- base + theme_prism(palette = "plasma")

p1 + p2
```

It is good to try matching `theme_prism()` palettes with their corresponding
`scale_colour_prism()` and/or `scale_fill_prism()` palettes.

```{r, fig.height=3.5}
# try out two more theme palettes and their corresponding colour palettes
p1 <- base + theme_prism(palette = "summer") + 
  scale_colour_prism(palette = "summer")
p2 <- base + theme_prism(palette = "stained_glass") + 
  scale_colour_prism(palette = "stained_glass")

p1 + p2
```

## Make your own ggprism theme

Many of these theme palettes are pretty ugly and not suitable for publication.
Even GraphPad themselves say that palettes such as `summer` and `mustard_fields`
["are less than appealing"](https://www.graphpad.com/support/faq/prism-color-schemes/).

Therefore, it would be good to be able to make your own themes based off of
`theme_prism()`. Fortunately, because `theme_prism()` is a 'complete' theme it
means that using it to make your own theme is super easy. You can read 
[this chapter](https://ggplot2-book.org/extensions.html) from the official 
ggplot2 book, or 
[this blog post](https://www.r-bloggers.com/2018/09/custom-themes-in-ggplot2/)
if you want to learn more of the specifics of making your own themes.

Otherwise, an example is given below.

```{r}
# define a new theme function based on the stained_glass palette
theme_new <- function(base_size = 14,
                      base_family = "sans",
                      base_fontface = "bold",
                      base_line_size = base_size / 14,
                      base_rect_size = base_size / 14,
                      axis_text_angle = 0,
                      border = FALSE) {
  theme_prism(palette = "stained_glass",
              base_size = base_size,
              base_family = base_family,
              base_fontface = base_fontface,
              base_line_size = base_line_size,
              base_rect_size = base_rect_size,
              axis_text_angle = axis_text_angle,
              border = border) %+replace% 
    theme(panel.background = element_rect(fill = "white",
                                          colour = NA),
          plot.background = element_rect(fill = "red",
                                         colour = NA),
          axis.line = element_line(colour = "black"),
          axis.ticks = element_line(colour = "black"))
}
```

```{r, fig.height=3.5}
# compare theme_prism() and our new theme function
p1 <- base + theme_prism()
p2 <- base + theme_new()

p1 + p2
```

